Estimating carrying capacity for juvenile salmon using quantile random forest models

### 2020-12-04 analysis/paper/paper.Rmd
Implemented revisions suggested by Mike, Sarah and Richie before final submission

---
title: "Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@merck.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::word_document2:
      fig_caption: yes
      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: false
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      keep_tex: true
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
abstract: |
  Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids. A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods may fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest (QRF) model to  provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale. QRF models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
keywords: |
  carrying capacity; Chinook salmon; quantile random forest; quantile regression; random forests
highlights: |
  These are the highlights. 
---

@@

@@

```{r sp-rec-comp}
# rename one population
spawn_rec_params %<>%
  mutate(Population = recode(Population,
                             'Chiwawa R.' = 'Chiwawa River'))

@@

keep_wtsds = c('Catherine Creek',
               'Chiwawa River',
               'Upper Grande Ronde River',
               'Hayden Creek',
               'Minam River',
               'Tucannon River',
               'Upper Lemhi',
               'Lostine River',
               'South Fork Salmon River')

@@

The decline of anadromous Pacific salmonids (*Oncorhynchus spp.*) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend. These actions are often categorized into four H’s: harvest modification, hatchery practices, hydrosystem operations, and habitat rehabilitation. Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality. Freshwater habitat capacity deficits have recently been identified as a major factor directly impacting population abundance which has been largely overlooked in Columbia Basin salmonids [@Bond2018; @Hinrichsen2020; @NOAA2020]. Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*O. mykiss*) in the Pacific Northwest, USA [@NOAA2016; @NOAA2016b]. Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles. However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge [@Bond2018; @Hinrichsen2020; @NOAA2020]. Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.

@@

The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River basin, Pacific Northwest, USA. Within the interior Columbia River basin two major runs of Chinook salmon occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics. Stream-type Chinook salmon adults enter freshwater from the ocean earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in freshwater before entering the ocean as smolts. Ocean-type Chinook salmon adults enter freshwater later (e.g., fall or winter) spawn lower in the watershed, and the juveniles may spend between several weeks and six months in freshwater before migrating to the ocean as subyearlings. Here, we focus on stream-type Chinook salmon, and in particular the juvenile summer rearing period during low flow, during which juveniles are often termed parr, referring to the camouflage markings that occur on their sides during this life-stage. Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU). The Upper Columbia spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River spring/summer-run is listed as threatened [@NOAA2016; @NOAA2016b]. Hereafter, we refer to both ESUs simply as Chinook salmon.

@@

## Study Site

@@

## Habitat Covariate Selection

@@

## QRF Model Fit

@@

## Extrapolating to Other Sites

To predict capacity at larger spatial scales, such as the watershed or population, we developed an extrapolation model based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of generalized random tessellation stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classification [@Whittier2011], the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions (parr/m) was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. The design weights are generated from how much of the watershed each site is meant to represent. Because the CHaMP sites were selected from strata that usually comprised unequal portions of that watershed, these weights must be accounted for to lead to unbiased model coefficients [@Nahorniak2015].  We fit two different extrapolation models, one that included watershed as a covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else. We then made predictions of linear capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart. These points do not represent specific segments of streams, however, so we needed to do some spatial averaging of capacity predictions to generate larger scale capacity estimates.

@@

## Model Validation

@@

## QRF Model

@@

## Extrapolating to Other Sites

@@

## A Tool to Estimate Habitat Capacity

@@

## Biological Expectations from QRF Model

@@

## Extrapolation Model

@@

## The Future: Improving Habitat Data

@@

## Conclusions and Next Steps

@@

```{r fish-hab-sites}
qrf_data %>%
  tabyl(Watershed) %>%
  adorn_totals() %>%
  adorn_pct_formatting() %>%
  rename(`n Sites` = n,
         Percent = percent) %>%
  kable(caption = 'The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer-run Chinook salmon parr capacity QRF model',
        booktabs = T,
        linesep = "") %>%
  kable_styling(font_size = 10,
                latex_options = c("hold_position"))

@@

```{r sr-table, eval = T}
spawn_rec_params %>%
  select(Population, data) %>%
  unnest(data) %>%
  group_by(Population) %>%
  summarise(`n Yrs` = n_distinct(BroodYear),
            Spawner_type = unique(Spawner_type)) %>%
  select(Population, `n Yrs`, Spawner_type) %>%
  mutate(Spawner_type = recode(Spawner_type,
                               'total spawners' = 'Spawners',
                               'redd counts' = 'Redds')) %>%
  rename(`Adult Data` = Spawner_type) %>%
  mutate(`Parr Data` = if_else(grepl('Chiwawa', Population),
                               'Parr Surveys',
                               'RST')) %>%
  full_join(spawn_rec_params %>%
              mutate(prnt_val = paste0(prettyNum(round(est), big.mark = ','), ' (', round(cv, 3), ')')) %>%
              select(Population, form, prnt_val) %>%
              mutate(form = recode(form,
                                   'BevertonHolt' = 'Beverton Holt',
                                   'Hockey' = 'Hockey Stick')) %>%
              mutate(form = factor(form,
                                   levels = c('Beverton Holt',
                                              'Ricker',
                                              'Hockey Stick',
                                              'QRF'))) %>%
              spread(form, prnt_val)) %>%
  filter(Population %in% keep_wtsds) %>%
  mutate(Population = recode(Population,
                             'Chiwawa R.' = 'Chiwawa River')) %>%
  kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile random forest (QRF) model. Numbers in parentheses are coefficients of variation.",
        booktabs = T,
        linesep = "") %>%
  kable_styling(font_size = 10,
                latex_options = c("hold_position")) %>%
  # kable_styling(latex_options = c('scale_down')) %>%
  column_spec(1, width = "1in") %>%
  column_spec(c(5:8), width = "0.5in")

@@

```{r wtsd-map, out.width = "85%", fig.cap = "Watersheds with CHaMP habitat data. Watersheds in black also contain paired fish data. Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork."}
knitr::include_graphics('../figures/ChampWtds.png')
```

>>>>

---
title: Estimating carrying capacity for juvenile salmon using quantile random forest models
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@merck.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::word_document2:
      fig_caption: yes
      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: false
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      keep_tex: true
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
abstract: |

  Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids (*Oncorhynchus* spp.). A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods may fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest model to  provide estimates of habitat carrying capacity for Chinook salmon (*O. tshawytscha*) parr during the summer months, at both the site and watershed scale. Quantile random forest models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on quantile random forest models, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
  **Key words:** carrying capacity; Chinook salmon; quantile random forest; quantile regression; random forests
keywords: |
  carrying capacity; Chinook salmon; quantile random forest; quantile regression; random forests
highlights: |
  These are the highlights. 
---

@@

# modify name of one watershed
hab_data %<>%
  mutate(Watershed = recode(Watershed,
                            "Yankee Fork" = "Yankee Fork Salmon"))

@@

```{r sp-rec-comp}
# rename one population
spawn_rec_params %<>%
  mutate(Population = recode(Population,
                             'Chiwawa R.' = 'Chiwawa River',
                             "Upper Lemhi" = "Upper Lemhi River"))

@@

keep_wtsds = c('Catherine Creek',
               'Chiwawa River',
               'Upper Grande Ronde River',
               'Hayden Creek',
               'Minam River',
               'Tucannon River',
               'Upper Lemhi River',
               'Lostine River',
               'South Fork Salmon River')

@@

The decline of anadromous Pacific salmon (*Oncorhynchus* spp.) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend. These actions are often categorized into four H’s: harvest modification, hatchery practices, hydrosystem operations, and habitat rehabilitation. Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality. Freshwater habitat capacity deficits have recently been identified as a major factor directly impacting population abundance which has been largely overlooked in Columbia Basin salmonids [@Bond2019; @Hinrichsen2020; @NOAA2020]. Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*O. mykiss*) in the Pacific Northwest, USA [@NOAA2016; @NOAA2016b]. Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles. However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge [@Bond2019; @Hinrichsen2020; @NOAA2020]. Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.

@@

The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River basin, Pacific Northwest, USA. Within the interior Columbia River basin two major runs of Chinook salmon occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics. Stream-type Chinook salmon adults enter freshwater from the ocean earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in freshwater before entering the ocean as smolts. Ocean-type Chinook salmon adults enter freshwater later (e.g., fall or winter), spawn lower in the watershed, and the juveniles may spend between several weeks and six months in freshwater before migrating to the ocean as subyearlings. Here, we focus on stream-type Chinook salmon, and in particular the juvenile summer rearing period during low flow, during which juveniles are often termed parr, referring to the camouflage markings that occur on their sides during this life-stage. Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU). The Upper Columbia spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River spring/summer-run is listed as threatened [@NOAA2016; @NOAA2016b]. Hereafter, we refer to both ESUs simply as Chinook salmon.

@@

## Study site

@@

## Habitat covariate selection

@@

## QRF model fit

@@

## Extrapolating to other Sites

To predict capacity at larger spatial scales, such as the watershed or population, we developed an extrapolation model based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of generalized random tessellation stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classification [@Whittier2011], the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions (parr/m) was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. The design weights are generated from how much of the watershed each site is meant to represent. Because the CHaMP sites were selected from strata that usually comprised unequal portions of that watershed, these weights must be accounted for to lead to unbiased model coefficients [@Nahorniak2015].  We fit two different extrapolation models, one that included watershed as a covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else. We then made predictions of linear capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart. These points do not represent specific segments of streams, however, so some form of spatial averaging of capacity predictions to generate larger scale capacity estimates is necessary.

@@

## Model validation

@@

## QRF model

@@

## Extrapolating to other sites

@@

## A tool to estimate habitat capacity

@@

## Biological expectations from QRF model

@@

## Extrapolation model

@@

## The future: improving habitat data

@@

## Conclusions and next steps

@@

```{r fish-hab-sites}
qrf_data %>%
  tabyl(Watershed) %>%
  adorn_totals() %>%
  adorn_pct_formatting() %>%
  rename(`n Sites` = n,
         Percent = percent) %>%
  kable(caption = 'The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer-run Chinook salmon parr capacity QRF model',
        booktabs = T,
        linesep = "") %>%
  kable_styling(font_size = 10,
                latex_options = c("hold_position"))
```

@@

```{r sr-table, eval = T}
spawn_rec_params %>%
  select(Population, data) %>%
  unnest(data) %>%
  group_by(Population) %>%
  summarise(`n Yrs` = n_distinct(BroodYear),
            Spawner_type = unique(Spawner_type)) %>%
  select(Population, `n Yrs`, Spawner_type) %>%
  mutate(Spawner_type = recode(Spawner_type,
                               'total spawners' = 'Spawners',
                               'redd counts' = 'Redds')) %>%
  rename(`Adult Data` = Spawner_type) %>%
  mutate(`Parr Data` = if_else(grepl('Chiwawa', Population),
                               'Parr Surveys',
                               'RST')) %>%
  full_join(spawn_rec_params %>%
              mutate(prnt_val = paste0(prettyNum(round(est), big.mark = ','), ' (', round(cv, 3), ')')) %>%
              select(Population, form, prnt_val) %>%
              mutate(form = recode(form,
                                   'BevertonHolt' = 'Beverton Holt',
                                   'Hockey' = 'Hockey Stick')) %>%
              mutate(form = factor(form,
                                   levels = c('Beverton Holt',
                                              'Ricker',
                                              'Hockey Stick',
                                              'QRF'))) %>%
              spread(form, prnt_val)) %>%
  filter(Population %in% keep_wtsds) %>%
  mutate(Population = recode(Population,
                             'Chiwawa R.' = 'Chiwawa River')) %>%
  kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile random forest (QRF) model. Numbers in parentheses are coefficients of variation.",
        booktabs = T,
        linesep = "") %>%
  kable_styling(font_size = 10,
                latex_options = c("hold_position")) %>%
  column_spec(1, width = "1.5in") %>%
  column_spec(c(5:8), width = "0.5in")

@@

```{r wtsd-map, out.width = "85%", fig.cap = "Watersheds with CHaMP habitat data. Watersheds in black also contain paired fish data. Watershed names are: 1 - Entiat, 2 - John Day, 3 - Lemhi, 4 - Methow, 5 - Minam, 6 - Secesh, 7 - South Fork Salmon, 8 - Tucannon, 9 - Upper Grande Ronde, 10 - Wenatchee, 11 - Yankee Fork Salmon."}
knitr::include_graphics('../figures/ChampWtds.png')
```
